TacticToe: Learning the Game of Tactics

### 2018-01-11 tactictoe_jv.tex
table of contents: for organizing

\section{Introduction}
%\begin{example} Proof automatically generated by \tactictoe for the user given 
%goal
%\small
%\begin{center}
%
%
%\begin{BVerbatim}[commandchars=\\\{\}]
%Goal: ``\(\forall\)l. FOLDL (\(\lambda\)xs x. SNOC x xs) [] l = l``
%Proof:
%  SNOC_INDUCT_TAC THENL
%  [ REWRITE_TAC [APPEND_NIL, FOLDL],
%    ASM_REWRITE_TAC [APPEND_SNOC, FOLDL_SNOC]
%      THEN CONV_TAC (DEPTH_CONV BETA_CONV)
%      THEN ASM_REWRITE_TAC [APPEND_SNOC, FOLDL_SNOC] ]
%\end{BVerbatim}
%
%
%\end{center}
%
%\end{example}
%

\todo{These 2 paragraphs are copied and slightly edited}
Many of the state-of-the-art interactive theorem provers (ITPs) such as 
  \holfour~\cite{hol4}, \hollight~\cite{Harrison09hollight}, 
  \isabelle~\cite{isabelle} 
  and \coq~\cite{coq-book}
  provide high-level parameterizable tactics for constructing the
  proofs.  Such tactics typically analyze the current goal state and
  assumptions, apply nontrivial proof transformations, which get
  expanded into possibly many basic kernel-level inferences or significant 
  parts of the proof term.
%\marginpar{not Coq}
  In this
  work we develop a tactic-level automation procedure for the \holfour ITP
  which guides selection of the tactics by learning from previous
  proofs.  Instead of relying on translation to first-order automated
  theorem provers (ATPs) as done by the hammer 
  systems~\cite{hammers4qed,tgck-cpp15}, the technique
%\marginpar{too many cites?}
%BlanchetteGKKU16,holyhammer,mizAR40
  directly searches for sequences of tactic applications that lead to
  the ITP proof, thus avoiding the translation and
  proof-reconstruction phases needed by the hammers.

@@

A first successful use of orthogonalization techniques for theorems appeared in
a work on lemma mining~\cite{ckju-jsc15} in \hollight. Theorems were mined from 
proofs recorded at the kernel level. Because the number of recorded lemmas was 
in millions, page ranking techniques was used as a filter to extract ``useful'' 
lemmas.

@@

\section{Tactic Abstraction}\label{sec:synthesis}
One of the great weakness of the previous version of \tactictoe is that
it could not create its own tactics. Indeed, sometimes no previous tactic is 
adapted for a certain goal and creating a new tactic is necessary.
%Experiments in section, show how different degrees of tactic synthesis could 
%improve the success rate of \tactictoe.
We present a way to create tactics with different arguments
by abstracting them and re-instantiating them using a predictor for that kind 
of argument. In the spirit of the orthogonalization method, we will try to 
create tactics that are more general but have the same effect as the 
ungeneralized one.

@@

The first necessary adaptation of the algorithm stems
from the fact that a proof is in general not a
path but a tree. This means that our search space has two branching factors: 
the choice of a tactic, and the number of goals produced by each tactic.
The first one can be thought as a disjunction of tactics. Indeed, 
if there exists a tactic that solves a goal then it is solved. And 
conversely, the second branching factor is a conjunction of goals. All pending 
goals must be proven in order to prove the parent goal. For example, in the 
case of induction tactics the pending goals are the base case and the inductive 
case.

@@

\begin{figure}{}
\begin{center}
\begin{tikzpicture}[node distance=0.7cm]
\node [tcircle] (0) {$g_0^i$};
\node [right of=0] (0r) {.\ .\ .};
\node [tcircle,right of=0r] (0rr) {$g_0^n$};
\node [left of=0] (0l) {.\ .\ .};
\node [tcircle,left of=0l] (0ll) {$g_0^1$};
\node [draw,fit=(0rr) (0ll)] {}; 
\node [above of=0,node distance=2cm] (2) {.\ .\ .};
\node [left of=2, tcircle] (2l) {$g_2^1$};
\node [tcircle,right of=2] (2r) {$g_2^p$};
\node [draw, fit=(2l) (2r)] (2b) {};

\node [tcircle, left of=2l, node distance=2cm] (1r) {$g_1^q$};
\node [left of=1r] (1) {.\ .\ .};
\node [tcircle, left of=1] (1l) {$g_1^1$};
\node [draw, fit=(1l) (1r)] (1b) {};

\node [right of=2r,node distance=2cm] (3l) {$\phantom{g_3^p}$};
\node [right of=3l] (3) {\phantom{.\ .\ .}};
\node [right of=3] (3r) {$\phantom{g_3^p}$};
\node [fit=(3l) (3r)] (3b) {};

\draw[-to,thick] (0) to node[xshift=-10](t1){$t_1$} (1b);
\draw[-to,thick] (0) to node[xshift=-10](t2){$t_2$} (2b);
\draw[-to,dotted,thick] (0) to (3b);
\end{tikzpicture}
\end{center}
\caption{Choice of a goal and a tactic}
\label{fig:choice}
\end{figure}

@@

\section{ATP and Lemma Prediction Integration}

General-purpose proof automation mechanisms which combine proof translation to
ATPs with machine learning (``hammers'') have become quite successful in
enhancing the automation level in proof assistants~\cite{hammers4qed}.
As external automated reasoning techniques sometimes outperform the combined 
power of tactics, we would like to combine the \tactictoe search with 
\holyhammer for \holfour~\cite{tgck-cpp15}. The principle is that we will 
create a \holyhammer tactic that can be used during the proof search. Since 
external calls to \holyhammer are computationally expensive (5 seconds at a 
minimum), we will also 
experiment with a ``small hammer'' tactic comprised of: a faster premise 
selection algorithm from previously proven theorems $\mathfrak{T}$ and
a short call to the internal prover \metis~\cite{metis}.

It is best tactic available but we do not have any data about it since it does 
not appear in any proofs. So as a first approximation, we 
artificially give it the best prediction scores on any goal and the best 
coverage.

\paragraph{Effect on orthogonalization of tactics}
During orthogonalization, the general purpose tactic will be the first 
competitor and if it manages to prove the goal it will subsume the original 
tactic on this goal since the empty set of goals is a subset of any set. 
Therefore, it will replace the original tactic in the feature vector that is 
competed for. This is desirable since we would like to remove alternative 
proofs.

\paragraph{Orthogonalization of theorems}
A ``hammer'' tactic can be useful to remove parallel
theorems. We consider a theorem to be parallel if it can be proven 
from previous theorems using a general tactic with a timeout equal or less 
than the one it is given during the proof search. Parallel theorems are marked 
and removed from the set of theorems $\mathfrak{T}$ from 
which premise selection is performed.

\begin{remark}
The ``smaller hammer'' is chosen to remove parallel theorems and as the first 
try during orthogonalization of tactics. 
\end{remark}

\paragraph{Effect on the proof search}
First, before the proof search, we preselect 500 theorems 
for the whole proof search tree using a premise selection algorithm 
that takes into account dependencies between theorems.
During the proof search, when a new goal is created or a fresh pending goal is 
considered, the ``hammer'' tactic is always tried first because if it succeeds 
the node would be closed rapidly.

\paragraph{Parallelization of external calls}
It is possible to parallelize calls to external provers so that they happen 
asynchronously during the proof search. We let the prover runs \tactictoe 
continue its proof in parallel of a ``hammer'' tactic call. Additional care 
needs to be taken after each proof step to close proofs solved by the general 
tactic. When the number of thread reaches a limit, a ``hammer'' tactic call 
should
wait for a thread to be freed but the proof search goes on and the feedback of 
the ``hammer'' tactic is delayed.

@@

\subsubsection{Tactic synthesis}
Here, we test the benefit of tactic synthesis compared to the old version.
In all those experiments orthogonalization is performed with radius 20 but not 
orthogonalization with respect to metis.

@@

\subsubsection{Hammer parameters}

The third experiment, presented in Table~\ref{tab:lh_param},
evaluates the effect of integrating the ``small hammer'' in the \tactictoe 
search.
At a first glance, the increased success rate is significant for all tested 
parameters. 
Further analysis reveals that increasing the number of premises from 8 to 16 
with a 
timeout of 0.02 seconds is detrimental. The $D_{19}$ experiment  demonstrates 
that 0.1 
seconds is a better time limit for reasoning with 16 premises. And the $D_{18}$ 
experiment reveals the disadvantage of unnecessarily increasing the timeout of 
\metis. This reduces the time available for the rest of the proof search, which 
makes the success rate drop.

@@

\pgfplotscreateplotcyclelist{my black}{
solid, mark repeat=100, mark phase=0, black!100\\
dashed, mark repeat=100, mark phase=0, black!100\\
solid, mark repeat=20, mark phase=10, black!100, mark=triangle*\\
solid, mark repeat=20, mark phase=10, black!75, mark=square*\\
solid, mark repeat=20, mark phase=10, black!50, mark=*\\
solid, mark repeat=20, mark phase=10, black!50, mark=o\\
}

\begin{figure}[p]
\centering           
\begin{tikzpicture}[scale=1]
\begin{axis}[
  legend style={anchor=south east, at={(0.9,0.1)}},
  width=\textwidth,
  height=0.7*\textwidth,xmin=0, xmax=10,
  ymin=0, ymax=4500,
  xtick={},
  ytick={},
  cycle list name=my black]
\addplot table[x=time, y=solved] {data/tt_ptime};
\addplot table[x=time, y=solved] {data/e_ptime};
\legend{\tactictoe,\eprover}
\end{axis}
\end{tikzpicture}
\caption{Number of problem solved in less than x seconds.}
\end{figure}

\begin{figure}[p]
\centering           
\begin{tikzpicture}[scale=1]
\begin{axis}[
  legend style={anchor=north east, at={(0.9,0.9)}},
  width=\textwidth,
  height=0.7*\textwidth,xmin=0, xmax=10,
  ymin=0, ymax=100,
  xtick={},
  ytick={},
  cycle list name=my black]
\addplot table[x=size, y=solved] {data/tt_percent};
\addplot table[x=size, y=solved] {data/e_percent};
\legend{\tactictoe,\eprover}
\end{axis}
\end{tikzpicture}
\caption{Percentage of problem solved with respect to the length of the 
original 
proof}
\end{figure}

\begin{table*}[t]
\centering\ra{1.3}
\small
\begin{tabular}{llllll}
\toprule
 ID & \multicolumn{2}{c}{nodes}   & \multicolumn{2}{c}{proof size} & 
 \multicolumn{1}{c}{time}\\ \cmidrule(lr){2-3} \cmidrule(lr){4-5} 
 \cmidrule(lr){6-6}
     & average & max & average & max & average \\
\midrule
 \tactictoe & 94.66 & 421 & 3.34 & 39 & 0.66 \\
\bottomrule
\end{tabular}
 \caption{\label{tab:stats} old search statistics}
\end{table*}

@@

>>>>

\section{Introduction}
%\begin{example} Proof automatically generated by \tactictoe for the user given 
%goal
%\small
%\begin{center}
%
%
%\begin{BVerbatim}[commandchars=\\\{\}]
%Goal: ``\(\forall\)l. FOLDL (\(\lambda\)xs x. SNOC x xs) [] l = l``
%Proof:
%  SNOC_INDUCT_TAC THENL
%  [ REWRITE_TAC [APPEND_NIL, FOLDL],
%    ASM_REWRITE_TAC [APPEND_SNOC, FOLDL_SNOC]
%      THEN CONV_TAC (DEPTH_CONV BETA_CONV)
%      THEN ASM_REWRITE_TAC [APPEND_SNOC, FOLDL_SNOC] ]
%\end{BVerbatim}
%
%
%\end{center}
%
%\end{example}
%
\vspace{100mm}

\tableofcontents

\vspace{100mm}
\todo{These 2 paragraphs are copied and slightly edited}
Many of the state-of-the-art interactive theorem provers (ITPs) such as 
  \holfour~\cite{hol4}, \hollight~\cite{Harrison09hollight}, 
  \isabelle~\cite{isabelle} 
  and \coq~\cite{coq-book}
  provide high-level parameterizable tactics for constructing the
  proofs.  Such tactics typically analyze the current goal state and
  assumptions, apply nontrivial proof transformations, which get
  expanded into possibly many basic kernel-level inferences or significant 
  parts of the proof term.
%\marginpar{not Coq}
  In this
  work we develop a tactic-level automation procedure for the \holfour ITP
  which guides selection of the tactics by learning from previous
  proofs.  Instead of relying on translation to first-order automated
  theorem provers (ATPs) as done by the hammer 
  systems~\cite{hammers4qed,tgck-cpp15}, the technique
%\marginpar{too many cites?}
%BlanchetteGKKU16,holyhammer,mizAR40
  directly searches for sequences of tactic applications that lead to
  the ITP proof, thus avoiding the translation and
  proof-reconstruction phases needed by the hammers.

@@

@@

\section{Abstraction}\label{sec:synthesis}
One of the great weakness of the previous version of \tactictoe is that
it could not create its own tactics. Indeed, sometimes no previous tactic is 
adapted for a certain goal and creating a new tactic is necessary.
%Experiments in section, show how different degrees of tactic synthesis could 
%improve the success rate of \tactictoe.
We present a way to create tactics with different arguments
by abstracting them and re-instantiating them using a predictor for that kind 
of argument. In the spirit of the orthogonalization method, we will try to 
create tactics that are more general but have the same effect as the 
ungeneralized one.

@@

(todo: missing link)
A proof is in general not a path but a tree. This means that our search space 
has two branching factors: 
the choice of a tactic, and the number of goals produced by each tactic.
The first one can be thought as a disjunction of tactics. Indeed, 
if there exists a tactic that solves a goal then it is solved. And 
conversely, the second branching factor is a conjunction of goals. All pending 
goals must be proven in order to prove the parent goal. For example, in the 
case of induction tactics the pending goals are the base case and the inductive 
case.

@@

\begin{figure}{}
\begin{center}
\begin{tikzpicture}[node distance=0.7cm]
\node [tcircle] (0) {$g_0^i$};
\node [right of=0] (0r) {.\ .\ .};
\node [tcircle,right of=0r] (0rr) {$g_0^n$};
\node [left of=0] (0l) {.\ .\ .};
\node [tcircle,left of=0l] (0ll) {$g_0^1$};
\node [draw,fit=(0rr) (0ll)] {}; 
\node [above of=0,node distance=2cm] (2) {.\ .\ .};
\node [left of=2, tcircle] (2l) {$g_j^1$};
\node [tcircle,right of=2] (2r) {$g_j^p$};
\node [draw, fit=(2l) (2r)] (2b) {};

\node [tcircle, left of=2l, node distance=3cm] (1r) {$g_1^q$};
\node [left of=1r] (1) {.\ .\ .};
\node [tcircle, left of=1] (1l) {$g_1^1$};
\node [draw, fit=(1l) (1r)] (1b) {};

\node [right of=2r,node distance=3cm] (3l) {$\phantom{g_3^p}$};
\node [right of=3l] (3) {\phantom{.\ .\ .}};
\node [right of=3] (3r) {$\phantom{g_3^p}$};
\node [fit=(3l) (3r)] (3b) {};

\node [fit=(1r) (2l)] (1m) {}; 
\node [fit=(2l) (3r)] (2m) {};

\draw[-to,thick] (0) to node[xshift=-10](t1){$t_1$} (1b);
\draw[-to,thick] (0) to node[xshift=-10](t2){$t_j$} (2b);
\draw[-to,dotted,thick] (0) to (1m);
\draw[-to,dotted,thick] (0) to (2m);
\draw[-to,dotted,thick] (0) to (3b);
\end{tikzpicture}
\end{center}
\caption{The two branching factors: choice of a goal $g_0^i$ and choice of a 
tactic $t_j$.}
\label{fig:choice}
\end{figure}

@@

\section{ATP Integration}

General-purpose proof automation mechanisms which combine proof translation to
ATPs with machine learning (``hammers'') have become quite successful in
enhancing the automation level in proof assistants~\cite{hammers4qed}.
As external automated reasoning techniques sometimes outperform the combined 
power of tactics, we would like to combine the \tactictoe search with 
\holyhammer for \holfour~\cite{tgck-cpp15}. The principle is that we will 
create a \holyhammer tactic that can be used during the proof search. Since 
external calls to \holyhammer are computationally expensive (5 seconds at a 
minimum), we will also 
experiment with a small hammer tactic comprised of: a faster premise 
selection algorithm from previously proven theorems $\mathfrak{T}$ and
a short call (0.1 seconds) to the internal prover \metis~\cite{metis} or a
longer call to external prover \eprover.

We construct this type of tactic manually but in the case of \metis it could be 
rediscovered automatically as it is essentially a call to \metis where its 
theorem list argument has been abstracted.

Since we do not always have the data to learn on which goal this general 
tactics are succesful, we assume as a first approximation that they are the 
strongest all around. And we try them first on each node during the proof 
search with the advantage that if it succeeds the node is closed rapidly in the 
case of \metis.

To integrate a general tactic that requires a higher time out such as \eprover, 
we make use of asynchronous calls to prevent the ATP 
from slowing down the search.  The number of parallel threads can be capped.
And if tactictoe or an asynchronous thread proves a goal, all descendants of 
these goal are deactivated terminating associated threads. This frees space for 
more pending threads associated to newly created nodes to start.

%needs to be taken after each proof step to close proofs solved by \eprover
%When the number of thread reaches a limit, a ``hammer'' tactic call 
%should
%wait for a thread to be freed but the proof search goes on and the feedback of 
%the ``hammer'' tactic is delayed. 
%
%
%It is best tactic available but we do not have any data about it since it does 
%not appear in any proofs. So , we 
%artificially try it on

%\paragraph{Effect on orthogonalization of tactics}
%During orthogonalization, the general purpose tactic will be the first 
%competitor and if it manages to prove the goal it will subsume the original 
%tactic on this goal since the empty set of goals is a subset of any set. 
%Therefore, it will replace the original tactic in the feature vector that is 
%competed for. This is desirable since we would like to remove alternative 
%proofs.

%\paragraph{Orthogonalization of theorems}
%A ``hammer'' tactic can be useful to remove parallel
%theorems. We consider a theorem to be parallel if it can be proven 
%from previous theorems using a general tactic with a timeout equal or less 
%than the one it is given during the proof search. Parallel theorems are marked 
%and removed from the set of theorems $\mathfrak{T}$ from 
%which premise selection is performed.
%
%\begin{remark}
%The ``smaller hammer'' is chosen to remove parallel theorems and as the first 
%try during orthogonalization of tactics. 
%\end{remark}

%\paragraph{Effect on the proof search}
%First, before the proof search, we preselect 500 theorems 
%for the whole proof search tree using a premise selection algorithm 
%that takes into account dependencies between theorems.
%During the proof search, when a new goal is created or a fresh pending goal is 
%considered, the ``hammer'' tactic is always tried first because if it succeeds 
%the node would be closed rapidly.

%\paragraph{Parallelization of external calls}
%It is possible to parallelize calls to external provers so that they happen 
%asynchronously during the proof search. We let the prover runs \tactictoe 
%continue its proof in parallel of a ``hammer'' tactic call.

@@

\subsubsection{Abstraction}
Here, we test the benefit of different kind of tactic abstraction. 
In all those experiments orthogonalization is performed with radius 20 but not 
orthogonalization with respect to metis.

@@

\subsubsection{ATP integration}

%The third experiment, presented in Table~\ref{tab:lh_param},
%evaluates the effect of integrating the ``small hammer'' in the \tactictoe 
%search.
%At a first glance, the increased success rate is significant for all tested 
%parameters. 
%Further analysis reveals that increasing the number of premises from 8 to 16 
%with a 
%timeout of 0.02 seconds is detrimental. The $D_{19}$ experiment  demonstrates 
%that 0.1 
%seconds is a better time limit for reasoning with 16 premises. And the 
%$D_{18}$ 
%experiment reveals the disadvantage of unnecessarily increasing the timeout of 
%\metis. This reduces the time available for the rest of the proof search, 
%which 
%makes the success rate drop.

The best setting for \metis was to be in previous experiments to be 16
theorems and

@@

Analysing failed proof attempt can give us an idea of the behaviour of the 
algorithm. In a proof search of 10 seconds, in average 500 (todo: real 
estimation) tactics are applied.
The productive tactics create an average of 100 (todo: real estimation) nodes. 
This means that in this settings 1 out of 5 tactics are productive.

Increasing the timeout to more than 10 seconds would probably be beneficial as 
we can see that the slope of the curve is far for being flat in Fig~..

\pgfplotscreateplotcyclelist{my black}{
solid, mark repeat=100, mark phase=0, black!100\\
dashed, mark repeat=100, mark phase=0, black!100\\
}

\begin{figure}[h]
\centering           
\begin{tikzpicture}[scale=1]
\begin{axis}[
  legend style={anchor=south east, at={(0.9,0.1)}},
  width=\textwidth,
  height=0.7*\textwidth,xmin=0, xmax=10,
  ymin=0, ymax=4500,
  xtick={},
  ytick={},
  cycle list name=my black]
\addplot table[x=time, y=solved] {data/tt_ptime};
\addplot table[x=time, y=solved] {data/e_ptime};
\legend{\tactictoe,\eprover}
\end{axis}
\end{tikzpicture}
\caption{Number of problem solved in less than x seconds.}
\end{figure}

(Todo: change next figure to an histogram)
\begin{figure}[h]
\centering           
\begin{tikzpicture}[scale=1]
\begin{axis}[
  legend style={anchor=north east, at={(0.9,0.9)}},
  width=\textwidth,
  height=0.7*\textwidth,xmin=0, xmax=10,
  ymin=0, ymax=100,
  xtick={},
  ytick={},
  cycle list name=my black]
\addplot table[x=size, y=solved] {data/tt_percent};
\addplot table[x=size, y=solved] {data/e_percent};
\legend{\tactictoe,\eprover}
\end{axis}
\end{tikzpicture}
\caption{Percentage of problem solved with respect to the length of the 
original 
proof}
\end{figure}

%\begin{table*}[t]
%\centering\ra{1.3}
%\small
%\begin{tabular}{llllll}
%\toprule
% ID & \multicolumn{2}{c}{nodes}   & \multicolumn{2}{c}{proof size} & 
% \multicolumn{1}{c}{time}\\ \cmidrule(lr){2-3} \cmidrule(lr){4-5} 
% \cmidrule(lr){6-6}
%     & average & max & average & max & average \\
%\midrule
% \tactictoe & 94.66 & 421 & 3.34 & 39 & 0.66 \\
%\bottomrule
%\end{tabular}
% \caption{\label{tab:stats} old search statistics}
%\end{table*}

@@

Future work:
  Promising techniques that could not be evaluated. Orthogonalization of 
  theorems~\cite{ckju-jsc15}. Term predictions.

Cite in future works.
%A first successful use of orthogonalization techniques for theorems appeared in
%a work on lemma mining~\cite{ckju-jsc15} in \hollight. Theorems were mined 
%from 
%proofs recorded at the kernel level. Because the number of recorded lemmas was 
%in millions, page ranking techniques was used as a filter to extract 
%``useful'' 
%lemmas.
